Wide-Field Effects in Redshifted 21~cm Power Spectra

### 2015-03-15 ms.tex
Modify figure 1, minor changes
From 25a1e5d:
============

Add new figure to git version control

In ms.tex,

(*) Modify figure 1
(*) Change text and caption accordingly
(*) Change double hyphens to single hyphens
(*) Make minor changes

\shorttitle{Wide--field Effects in EoR Power Spectra}
\shortauthors{Thyagarajan et~al.}

@@

\title{Wide--Field Effects in Redshifted 21~cm Power Spectra}

@@

Foreground emission is currently the primary limitation to detection of redshifted H{\sc i} emission from the epoch of reionization. Modern radio telescopes that target this cosmological signal are typically wide--field instruments. Through modeling of delay spectra measured between antenna pairs, it has recently emerged that wide--field measurements imprint a characteristic {\it pitchfork}--shaped signature in this Fourier domain. This predicted feature is characterized by enhanced power from foreground emission mapped to regions near the horizon and plays a significant role in determining the contamination of the cosmological H{\sc i} signal. With MWA data sensitivity improved by coherently averaging 12 independent snapshots aligned in local sidereal time across different observing nights, we confirm the prediction with a signal--noise ratio $>10$.

@@

The epoch of reionization (EoR) commenced following the formation of the first stars and galaxies. It is characterized by a period of non--linear growth of matter density perturbations and astrophysical evolution in the Universe's history. Detection of redshifted 21~cm radiation of H{\sc i} from this epoch is one of the most promising probes of the non--linear evolution of large scale structure characteristic of this epoch \citep{sun72,sco90,mad97,toz00,ili02}.

@@

A recent study by \citet{thy15} using instrument and all--sky foreground models has shed new light on the significant role played by the interplay between foreground emission and the nature of wide--field measurements, typical of most modern EoR experiments. This paper presents a follow--up analysis of that study. By using a larger volume of MWA data obtained independently over different nights pointed at the same region of sky, we confirm with high significance the presence of certain key characteristics (referred to as the {\it pitchfork} signature) of wide--field measurements predicted in the preceding study.

\S\ref{sec:wide-field} gives an overview of the role of wide--field measurements in the delay spectral domain and the predicted {\it pitchfork} signature. In \S\ref{sec:MWA}, we describe the data analysis procedure used to improve the dynamic range of delay spectra of MWA data. We describe the results and confirm the presence of the predicted wide--field effects in \S\ref{sec:results}. \S\ref{sec:summary} summarizes our findings.

\section{Wide--Field Effects in Delay Spectrum}\label{sec:wide-field}

\citet{thy15} have described in detail the effects of wide--field measurements as seen in the delay spectra of interferometer {\it visibilities}. Here, we give a brief overview of the wide--field signature predicted therein.

@@

In wide--field measurements, the steep rise in subtended solid angle near the horizon for a fixed delay bin size significantly enhances the integrated emission near the horizon delay limits. This is found to be true for diffuse emission even on wide antenna spacings because their foreshortening towards the horizon makes them sensitive to large angular scales that match the inverse of their foreshortened lengths.

Typically, an antenna is most sensitive towards its primary field of view relative to the rest of the sky, which in conjunction with the aforementioned wide--field effects results in a characteristic {\it pitchfork} signature in the delay spectrum.

@@

Figure~\ref{fig:avg-fhd-delay-spectra} shows the delay spectra obtained by averaging LST aligned delay spectra from individual snapshots from different nights. The dynamic range (in power spectrum units) in the averaged data is $\gtrsim 10$ higher compared to that in the {\it zenith} snapshot used in \citet{thy15}, and is consistent with the improvement expected in averaging 12 independent snapshots. With this improvement in sensitivity, the foreground power near the horizon limits (white dotted lines) has become $\gtrsim 10$ times more prominent. We also note that faint horizontal features appear at $\tau = \pm 0.78\,\mu$s also as a result of lowering thermal fluctuations which were not seen earlier in the individual snapshots presented in \citet{thy15}, thus confirming effective lowering of thermal fluctuations. We identify these faint features as the response in delay space of the MWA coarse band edges flagged periodically every 1.28~MHz.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{multi_baseline_CLEAN_fhd_avg_visibilities_amplitudes_185.0_MHz_30.7_MHz.eps}
\caption{Amplitudes of delay power spectra obtained by averaging 12 snapshots of LST aligned MWA data. The $x$--axis, denoted by $|\boldsymbol{b}|$ (and $k_\perp$), represents angular (and spatial) scales in the plane of the sky while the $y$--axis, shown in $\tau$ and $k_\parallel$, denotes the spatial scales along the line of sight. White dotted lines are the horizon delay limits. Power near the horizon limits caused by wide--field effects are prominent. Faint horizontal features at $\tau=\pm 0.78\,\mu$s are visible due to effective lowering of thermal fluctuations and are the response to periodic coarse band edge flagging of MWA data every 1.28~MHz. \label{fig:avg-fhd-delay-spectra}}
\end{figure}

Figure~\ref{fig:3-baseline-comparison-delay-spectra} shows the amplitudes of averaged delay spectra on three selected baseline vectors oriented northward. Data and noiseless simulations (using foreground and instrument models described in \citet{thy15}) are shown in black and red respectively. The horizontal dotted black line denotes {\it rms} of thermal fluctuations estimated from data. The vertical dashed line denotes horizon delay limits, and the vertical dot--dashed lines denote delays at which the responses to coarse band edge flagging are expected.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{3_baseline_comparison_CLEAN_fhd_avg_visibilities_amplitudes_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra amplitudes on three antenna spacings oriented northward obtained by coherent averaging of 12 snapshots aligned in LST. The data and models are shown in black and red respectively. The antenna spacings are 83.5~m ({\it top}), 104.2~m ({\it middle}), and 171~m ({\it bottom}). The horizontal dotted line is the {\it rms} of thermal fluctuations. The vertical dashed lines denote the horizon delay limits. The vertical dot--dashed lines at $\tau=\pm 0.78\,\mu$s correspond to grating responses of periodic flagging of bandpass at intervals of 1.28~MHz. The dynamic range has increased by a factor $\sim 10$ as a result of coherent averaging relative to that in \citet{thy15}. The peaks close to the horizon delay limits are distinctly visible at $\sim 10$--1000~$\sigma$ levels. Differences between model and data are primarily attributed to uncertainties in the foreground model and the MWA tile power pattern. \label{fig:3-baseline-comparison-delay-spectra}}
\end{figure}

@@

We focus on the peaks near the horizon limits in the data. Typically, the power near the negative horizon limit is seen with a signal--noise ratio (SNR) $\sim$~10--100, while that around the positive horizon limit is $\sim$~100--1000. The models of \citet{thy15} have noted that the foreground power near the horizon limits is due to the nature of wide--field measurements, and is predominantly composed of diffuse emission especially on baseline lengths $\lesssim 100$~m. Based on the morphological agreement between the data and the models, we conclude the features noted in our current analysis are a robust detection of the {\it pitchfork} signature predicted in \citet{thy15}.

@@

In a recent study, through the use of models for an all--sky foreground and the instrument, the role of wide--field nature of modern radio interferometric measurements, hitherto unpredicted, has emerged. It is referred to as the {\it pitchfork} signature in the delay spectral domain, and is characterized by significant foreground power in the primary field of view and an enhancement in power near the horizon. The latter is due to the highly non--linear mapping between geometric delays and subtended solid angles besides an increase in sensitivity to larger scales caused by foreshortening of baseline lengths along these directions. This feature could not be confirmed in individual MWA snapshot observations due to the high level of thermal fluctuations.

By averaging 12 independent snapshots tightly aligned in LST from MWA observations, we improve the dynamic range in delay power spectra by a factor $\gtrsim$~10 while maintaining the coherency of the observed {\it visibilities}. Foreground power in delay bins near the horizon is distinctly visible with a signal--noise ratio $>10$. In addition, there is a close resemblance in the morphologies of delay power spectra from data and models. Together, they allow us to make a robust confirmation of the nature of wide--field measurements in EoR experiments.

@@

This work was supported by the U. S. National Science Foundation (NSF) through award AST--1109257. DCJ is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST--1401708. JCP is supported by an NSF Astronomy and Astrophysics Fellowship under award AST-1302774. This work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO. We acknowledge the Wajarri Yamatji people as the traditional owners of the Observatory site. Support for the MWA comes from the NSF (awards: AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S. Air Force Office of Scientific Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020). Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant). The Australian Federal government provides additional support via the Commonwealth Scientific and Industrial Research Organisation (CSIRO), National Collaborative Research Infrastructure Strategy, Education Investment Fund, and the Australia India Strategic Research Fund, and Astronomy Australia Limited, under contract to Curtin University. We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government.

>>>>

\shorttitle{Wide-field Effects in EoR Power Spectra}
\shortauthors{Thyagarajan et~al.}

@@

\title{Wide-Field Effects in Redshifted 21~cm Power Spectra}

@@

Foreground emission is currently the primary limitation to detection of redshifted H{\sc i} emission from the epoch of reionization. Modern radio telescopes that target this cosmological signal are typically wide-field instruments. Through modeling of delay spectra measured between antenna pairs, it has recently emerged that wide-field measurements imprint a characteristic {\it pitchfork}-shaped signature in this Fourier domain. This predicted feature is characterized by enhanced power from foreground emission mapped to regions near the horizon and plays a significant role in determining the contamination of the cosmological H{\sc i} signal. With MWA data sensitivity improved by coherently averaging 12 independent snapshots aligned in local sidereal time across different observing nights, we confirm the prediction with a signal-noise ratio $>10$.

@@

The epoch of reionization (EoR) commenced following the formation of the first stars and galaxies. It is characterized by a period of non-linear growth of matter density perturbations and astrophysical evolution in the Universe's history. Detection of redshifted 21~cm radiation of H{\sc i} from this epoch is one of the most promising probes of the non-linear evolution of large scale structure characteristic of this epoch \citep{sun72,sco90,mad97,toz00,ili02}.

@@

A recent study by \citet{thy15} using instrument and all-sky foreground models has shed new light on the significant role played by the interplay between foreground emission and the nature of wide-field measurements, typical of most modern EoR experiments. This paper presents a follow-up analysis of that study. By using a larger volume of MWA data obtained independently over different nights pointed at the same region of sky, we confirm with high significance the presence of certain key characteristics (referred to as the {\it pitchfork} signature) of wide-field measurements predicted in the preceding study.

\S\ref{sec:wide-field} gives an overview of the role of wide-field measurements in the delay spectral domain and the predicted {\it pitchfork} signature. In \S\ref{sec:MWA}, we describe the data analysis procedure used to improve the dynamic range of delay spectra of MWA data. We describe the results and confirm the presence of the predicted wide-field effects in \S\ref{sec:results}. \S\ref{sec:summary} summarizes our findings.

\section{Wide-Field Effects in Delay Spectrum}\label{sec:wide-field}

\citet{thy15} have described in detail the effects of wide-field measurements as seen in the delay spectra of interferometer {\it visibilities}. Here, we give a brief overview of the wide-field signature predicted therein.

@@

In wide-field measurements, the steep rise in subtended solid angle near the horizon for a fixed delay bin size significantly enhances the integrated emission near the horizon delay limits. This is found to be true for diffuse emission even on wide antenna spacings because their foreshortening towards the horizon makes them sensitive to large angular scales that match the inverse of their foreshortened lengths.

Typically, an antenna is most sensitive towards its primary field of view relative to the rest of the sky, which in conjunction with the aforementioned wide-field effects results in a characteristic {\it pitchfork} signature in the delay spectrum.

@@

Figure~\ref{fig:delay-spectra} shows the delay spectra obtained from a single snapshot of MWA data (top), averaging LST aligned delay spectra from 12 individual snapshots from MWA observations on different nights (middle), and from modeling \citep{thy15} with no thermal noise added (bottom), for reference. The dynamic range (in power spectrum units) in the averaged data (middle) is a factor $\gtrsim 10$ higher relative to that in a single snapshot (top) similar to the one used in \citet{thy15}, and is consistent with the improvement expected from averaging 12 independent snapshots. With this improvement in sensitivity, the foreground power near the horizon limits (white dotted lines) has become $\gtrsim 10$ times more prominent. We now note that faint horizontal features appear at $\tau = \pm 0.78\,\mu$s also as a result of lowering thermal fluctuations which is not seen in the single snapshot, thus confirming effective lowering of thermal fluctuations. We identify these faint features as the response in delay space of the MWA coarse band edges flagged periodically every 1.28~MHz.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{multi_baseline_CLEAN_fhd_avg_visibilities_amplitudes_185.0_MHz_30.7_MHz.eps}
\includegraphics[width=\linewidth]{multi_baseline_fhd_sim_visibilities_amplitudes_comparison_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra obtained from a single snapshot ({\it top}), by averaging 12 snapshots of LST aligned MWA data ({\it middle}), and from modeling with no thermal noise added ({\it bottom}) shown for reference. The $x$-axis, denoted by $|\boldsymbol{b}|$ (and $k_\perp$), represents angular (and spatial) scales in the plane of the sky while the $y$-axis, shown in $\tau$ and $k_\parallel$, denotes the spatial scales along the line of sight. White dotted lines are the horizon delay limits. Dynamic range in the delay power spectra of MWA data has increased by a factor $\sim 10$ after averaging (middle) relative to that in a single snapshot (top). Power near the horizon limits caused by wide-field effects are prominent. Faint horizontal features at $\tau=\pm 0.78\,\mu$s are visible due to effective lowering of thermal fluctuations and are the response to periodic coarse band edge flagging of MWA data every 1.28~MHz. \label{fig:delay-spectra}}
\end{figure}

Figure~\ref{fig:3-baseline-comparison-delay-spectra} shows the amplitudes of averaged delay spectra on three selected baseline vectors oriented northward. Data and noiseless models are shown in black and red respectively. The horizontal dotted black line denotes {\it rms} of thermal fluctuations estimated from data. The vertical dashed line denotes horizon delay limits, and the vertical dot-dashed lines denote delays at which the responses to coarse band edge flagging are expected.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{3_baseline_comparison_CLEAN_fhd_avg_visibilities_amplitudes_185.0_MHz_30.7_MHz.eps}
\caption{Delay power spectra on three antenna spacings oriented northward, obtained by coherent averaging of 12 snapshots aligned in LST. The averaged data and models are shown in black and red respectively. The antenna spacings are 83.5~m ({\it top}), 104.2~m ({\it middle}), and 171~m ({\it bottom}). The horizontal dotted line is the {\it rms} of thermal fluctuations. The vertical dashed lines denote the horizon delay limits. The vertical dot-dashed lines at $\tau=\pm 0.78\,\mu$s correspond to grating responses of periodic flagging of bandpass at intervals of 1.28~MHz. The peaks close to the horizon delay limits are distinctly visible at $\sim 10$-1000~$\sigma$ levels. Differences between model and data are primarily attributed to uncertainties in the foreground model and the MWA tile power pattern. \label{fig:3-baseline-comparison-delay-spectra}}
\end{figure}

@@

We focus on the peaks near the horizon limits in the data. Typically, the power near the negative horizon limit is seen with a signal-noise ratio (SNR) $\sim$~10-100, while that around the positive horizon limit is $\sim$~100-1000. The models of \citet{thy15} have noted that the foreground power near the horizon limits is due to the nature of wide-field measurements, and is predominantly composed of diffuse emission especially on baseline lengths $\lesssim 100$~m. Based on the morphological agreement between the data and the models, we conclude the features noted in our current analysis are a robust detection of the {\it pitchfork} signature predicted in \citet{thy15}.

@@

In a recent study, through the use of models for an all-sky foreground and the instrument, the role of wide-field nature of modern radio interferometric measurements, hitherto unpredicted, has emerged. It is referred to as the {\it pitchfork} signature in the delay spectral domain, and is characterized by significant foreground power in the primary field of view and an enhancement in power near the horizon. The latter is due to the highly non-linear mapping between geometric delays and subtended solid angles besides an increase in sensitivity to larger scales caused by foreshortening of baseline lengths along these directions. This feature could not be confirmed in individual MWA snapshot observations due to the high level of thermal fluctuations.

By averaging 12 independent snapshots tightly aligned in LST from MWA observations, we improve the dynamic range in delay power spectra by a factor $\gtrsim$~10 while maintaining the coherency of foreground contributions in the observed {\it visibilities}. Foreground power in delay bins near the horizon is distinctly visible with a signal-noise ratio $>10$. In addition, there is a close resemblance in the morphologies of delay power spectra from data and models. Together, they allow us to make a robust confirmation of the nature of wide-field measurements in EoR experiments.

@@

This work was supported by the U. S. National Science Foundation (NSF) through award AST-1109257. DCJ is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1401708. JCP is supported by an NSF Astronomy and Astrophysics Fellowship under award AST-1302774. This work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO. We acknowledge the Wajarri Yamatji people as the traditional owners of the Observatory site. Support for the MWA comes from the NSF (awards: AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S. Air Force Office of Scientific Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020). Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant). The Australian Federal government provides additional support via the Commonwealth Scientific and Industrial Research Organisation (CSIRO), National Collaborative Research Infrastructure Strategy, Education Investment Fund, and the Australia India Strategic Research Fund, and Astronomy Australia Limited, under contract to Curtin University. We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government.
